Twenty years of dynamic occupancy models: a review of applications and a look towards the future

### 2024-06-11 Manuscript/domReview.qmd
Changes to introduction
Added more general introduction to species distributions

embed-resources: true
bibliography: references.bib
---

@@

@@

DOMs display great variation of scale in all senses. Analyses in our samples range from the hyper-local at individual parks [@zúñiga-vega2019] to continental-scale studies \[@zuckerberg2011\]. Similar diversity in study duration and site quantity indicates further emphasises the breadth of studies which DOMs may be fit to. Method of species detection also need not be a limitation for the implementation of DOMs, as articles in our sample show that many approaches may be coerced into data suitable for this model class. While studies pre-designed to match hierarchical data inputs may be simplest to align with model assumptions, post-hoc manipulation of other data such as long-term citizen monitoring programs \[@zuckerberg2011; @peach2019\], disparate agency modelling programs [@mcgowan2020], or camera trapping data [@davis2018] is feasible with careful consideration. Detections from multiple sources, with associated variation in efficiency, have also been incorporated by authors [@warrier2020; @pitman2017].

>>>>

bibliography: references.bib
---

@@

Ecologists have long sought to understand the processes by which species are distributed across landscapes and throughout time \[@humboldt1849\]. Research on these patterns of occurrence is valuable in many ways — simply describing where a species occurs may be of use in monitoring programs, and identifying the drivers of distribution can help to determine how best to manage or conserve a species.

In recent decades, interest in distributions has precipitated numerous methods for quantitatively modelling species distributions \[@elith2009\]. In their typical form species distribution models (SDMs) correlate species occurrence records with environmental covariates to describe where species occur on the landscape. They have been implemented in several thousand scientific articles and are frequently relied upon to support management and conservation programs. Despite their popularity SDMs do have considerable downsides. Typically, they do not explicitly account for imperfect detection, where a species is present at a site but is not detected. Imperfect detection is known to lead to biased estimates of occupancy and is a ubiquitous and near-unavoidable feature of wildlife presence-absence data. SDMs also tend to be most effective when species are at equilibrium states, and can struggle to predict to novel environments or future conditions.

Another approach to modelling species occurrence is the occupancy model, first introduced in @mackenzie2002. Like SDMs, occupancy models estimate probabilities of species occurrence; unlike SDMs, occupancy models account for imperfect detection by explicitly modelling the process of detection. @mackenzie2003 extends the original occupancy model to describe change in species occurrence across multiple time-steps. Despite the similarity in their output, SDMs and occupancy models have developed largely distinctly.

@@

DOMs display great variation of scale in all senses. Analyses in our samples range from the hyper-local at individual parks [@zúñiga-vega2019] to continental-scale studies [@zuckerberg2011]. Similar diversity in study duration and site quantity indicates further emphasises the breadth of studies which DOMs may be fit to. Method of species detection also need not be a limitation for the implementation of DOMs, as articles in our sample show that many approaches may be coerced into data suitable for this model class. While studies pre-designed to match hierarchical data inputs may be simplest to align with model assumptions, post-hoc manipulation of other data such as long-term citizen monitoring programs [@zuckerberg2011; @peach2019], disparate agency modelling programs [@mcgowan2020], or camera trapping data [@davis2018] is feasible with careful consideration. Detections from multiple sources, with associated variation in efficiency, have also been incorporated by authors [@warrier2020; @pitman2017].
